Fused regression for network inference in multiple species

### 2015-05-14 paper1.tex
hi

\title{Fused regression for network inference in multiple species}
\author{Kari Y. Lam, Zachary M. Westrick, Christian Mueller, Lionel Christiaen, Richard Bonneau}
\maketitle

@@

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become more attainable. Already, there is enough data in some model organisms for the development of high quality networks; however, this is still not true of many other model systems. Luckily, the problem of inferring gene regulatory networks is not an independent problem. There has been a great deal of work showing that functional conservation exists in gene regulatory networks even across large evolutionary distance. A high quality network in one species should be a useful source when learning the network of a related species. But most existing inference algorithms do not incorporate this information.

We employ a linear model of gene expression in which the rate of expression is assumed to be proportional to a weighted sum of transcription factor expression. These weights represent the regulatory interactions we wish to learn. We can fit these weights simultaneously in multiple species of interest using time series and steady state expression data. This is similar to many existing network inference methods and produces identical answers to fitting each species separately. We introduce additional constraints on the problem that penalize differences between regulatory weights which are expected to be similar. The choice of which interactions to fuse together follows from the observation that orthologous genes are likely to have similar regulatory interactions.

A few approaches to multi-source network inference have been proposed; however, current approaches either require one-to-one orthology [aviv regev paper], which is useful for analyzing cell lines but not for most cross species comparisons, or restrict analysis to the subset of genes with orthologs in every network [ibm paper]. Our approach allows simultaneous inference of the whole gene regulatory networks of species that may have partial orthology overlap. We demonstrate performance gains in networks learned from simulated data and real data from B. subtilis and B. anthracis, where a large fraction of genes have no orthologs. Though this method was developed for multi-species network inference, it is also useful in the easier problem of learning networks in multiple cell lines.

@@

\section{Methods}
\subsection{GRN}
We begin with an existing framework for estimating the interaction weights in a gene regulatory network (GRN). In the Inferelator algorithm, linear differential equations are used to model changes in gene expression. The rate at which the expression of gene $i$, $x_i$, changes is governed by degradation of existing transcripts with rate $\alpha$ plus a linear combination of transcription factor (TF) expressions.
\begin{equation}
\frac{\mathrm d}{\mathrm d t} x_i = -\alpha_{i}x_{i} + \sum \beta_{i,j}x_{j}
\end{equation}
We are interested in learning $\beta$, the matrix defining the influence of each TF on each gene. We fix the decay rate $\alpha$ for all genes, and set it assuming a time-constant of 10 minutes [CITATIONS], as in [CITATIONS]. Let $x_i(t)$ be the expression of gene $i$ at time $t$. Given time-series data on the expression of gene $i$ at timepoints $t_k$ and $t_{k+1}$, we can approximate the rate of change of $x_i$ as $x_i'(t_k)=\frac{x_i(t_{k+1})-x_i(t_k)}{t_{k+1}-t_k}$. We can treat steady-state data as having a derivative of zero. This gives us, for each gene $i$ and time $t_{k}$ an equation

@@

\subsection{Fused GRN}
Related species should be governed by similar but not necessarily identical gene regulatory networks. We approach the problem of cross-species GRN inference by introducing constraints into the above regression formulation to penalize differences between interaction weights that are expected to be similar based on gene orthology. We can then solve the penalized regression problems simultaneously, in order to obtain a GRN for each species. Consider the case of organisms $A$ and $B$, governed by GRNs $\beta^A$ and $\beta^B$ (the following approach applies equally well to more than two species but for simplicity we continue with the case of two species). If TF $g^A$ in organism $A$ and TF $h^B$ in organism $B$ are orthologs, and gene $k^A$ and $l^B$ are orthologs, then we expect that the $g^A \rightarrow k^A$ interaction weight should be similar to the $h^B \rightarrow l^B$ interaction weight, and we introduce a fusion constraint between these analogous interactions. In terms of the above regression formulation, we expect that $\beta^A_{g,k} \approx \beta^B_{h,l}$, and include a penalty term $\lambda_S(\beta^A_{g,k} - \beta^B_{h,l})^2$ to the quantity being minimized in order to ensure similarity. The parameter $\lambda_S$ controls how highly to penalize differences between fused coefficients. This gives us the final equation to be minimized 
\begin{equation}
\argmin_{(\beta^A, \beta^B)} \displaystyle\sum_{S \in (A, B)} \vert \vert X^S\beta^S - Y^S \vert \vert ^2 + \lambda_R \vert \vert \beta^S \vert \vert ^2 + \displaystyle \sum_{\beta^{S_1}_{g,k} \approx \beta^{S_2}_{h,l}} \lambda_S(\beta^{S_1}_{g,k} - \beta^{S_2}_{h,l})^2
\end{equation}

@@

[DIGRESSION?]

When there is a one-to-one orthology between the species being considered (ie different cell-lines of the same organism), the choice of $\lambda_S$ allows one to interpolate between fitting each network independently ($\lambda_S=0$) and pooling data together as if it came from one source ($\lambda_S=\inf$). Part of the appeal of the approach, however, is that it allows pooling of data even when there is incomplete orthology. By introducing constraints on the similarity of individual interactions, rather than on the networks as a whole [CITE IBM?], we can pool some information across species even when an arbitrarily small fraction of genes have orthologs. This is particularly useful when dealing with a large number of species; pairwise orthologies may be nearly complete even when the number of genes present in every organism is small. 
\subsubsection{Equivalent prior}
I want to talk about the fusion constraints as a Gaussian prior, and talk about the adjustment stuff

>>>>

\title{Fused regression for network inference in multiple species}
\author{Kari Y. Lam, Zachary M. Westrick, Christian L. M\"{u}ller, Lionel Christiaen, Richard Bonneau}
\maketitle

@@

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become more attainable. Already, there is enough data in some model organisms for the development of high quality networks; however, this is still not true of many other model systems. Luckily, the problems of inferring gene regulatory networks in multiple species are not independent, due to the presence of conserved structure. There has been a great deal of work showing that functional conservation exists in gene regulatory networks even across large evolutionary distance. Abundant data in one species should be a useful source when learning the network of a related species. But most existing inference algorithms do not incorporate this information.

We employ a linear model of gene expression in which the rate of expression is assumed to be proportional to a weighted sum of transcription factor expression. These weights represent the regulatory interactions we wish to learn. We can fit these weights simultaneously in multiple species of interest using time series and steady state expression data. This is similar to many existing network inference methods and produces identical answers to fitting each species separately. To take advantage of cross-species information, we introduce additional constraints on the problem that penalize differences between regulatory weights which are expected to be similar. The choice of which interactions to fuse together follows from the observation that orthologous genes are likely to have similar regulatory interactions.

A few approaches to multi-source network inference have been proposed; however, current approaches either require that the sources have identical sets of genes [aviv regev paper], which is useful for analyzing cell lines but not for most cross species comparisons, or restrict analysis to the subset of genes with orthologs in every species [ibm paper], which limits the applicability of the model. Our approach allows simultaneous inference of the whole gene regulatory networks of species that may have partial orthology overlap. We demonstrate performance gains in networks learned from simulated data and real data from Bacillus subtilis and Bacillus anthracis, where a large fraction of genes have no orthologs. Though this method was developed for multi-species network inference, it is also useful in the easier problem of learning networks in multiple cell lines.

@@

\section{Methods}
\subsection{Gene regulatory network}
We represent gene regulatory networks as a matrix of weights, where rows index transcription factors and columns index genes. The weight in a given position represents the regulatory network on that TF on that gene; positive weights represent activation, negative weights represent repression, and 0 weights represent the absence of an interaction. We begin with an existing framework for estimating the interaction weights in a gene regulatory network (GRN). In the Inferelator algorithm, linear differential equations are used to model changes in gene expression. The rate at which the observed mRNA expression of gene $i$, $x_i$, changes is governed by degradation of existing transcripts with rate $\alpha$ plus a linear combination of transcription factor (TF) expressions. 
\begin{equation}
\frac{\mathrm d}{\mathrm d t} x_i = -\alpha_{i}x_{i} + \sum \beta_{i,j}x_{j}
\end{equation}
We are interested in learning $\beta$, the matrix defining the influence of each TF on each gene. We fix the decay rate $\alpha$ for all genes, and set it assuming a time-constant of 10 minutes [CITATIONS], as in [CITATIONS]. Let $x_i(t)$ be the expression of gene $i$ at time $t$. Given time-series data on the expression of gene $i$ at timepoints $t_k$ and $t_{k+1}$, we can approximate the rate of change of $x_i$ as $x_i'(t_k)=\frac{x_i(t_{k+1})-x_i(t_k)}{t_{k+1}-t_k}$. We can treat steady-state data as having a derivative of zero. This gives us, for each gene $i$ and time $t_{k}$ an equation

@@

\subsection{Fused gene regulatory networks}
We hypothesize that related species should be governed by similar but not necessarily identical gene regulatory networks. We approach the problem of cross-species GRN inference by introducing constraints into the above regression formulation to penalize differences between interaction weights that are expected to be similar based on gene orthology. We can then solve the penalized regression problems simultaneously, in order to obtain a GRN for each species. Consider the case of organisms $A$ and $B$, governed by GRNs $\beta^A$ and $\beta^B$ (the following approach applies equally well to more than two species but for simplicity we continue with the case of two species). If TF $g^A$ in organism $A$ and TF $h^B$ in organism $B$ are orthologs, and gene $k^A$ and $l^B$ are orthologs, then we expect that the $g^A \rightarrow k^A$ interaction weight should be similar to the $h^B \rightarrow l^B$ interaction weight, and we introduce a fusion constraint between these analogous interactions. In terms of the above regression formulation, we expect that $\beta^A_{g,k} \approx \beta^B_{h,l}$, and include a penalty term $\lambda_S(\beta^A_{g,k} - \beta^B_{h,l})^2$ to the quantity being minimized in order to ensure similarity. The parameter $\lambda_S$ controls how highly to penalize differences between fused coefficients and controls the tradeoff between fitting the expression data and producing a set of networks that conform to evolutionary constraints. This gives us the final equation to be minimized 
\begin{equation}
\argmin_{(\beta^A, \beta^B)} \displaystyle\sum_{S \in (A, B)} \vert \vert X^S\beta^S - Y^S \vert \vert ^2 + \lambda_R \vert \vert \beta^S \vert \vert ^2 + \displaystyle \sum_{\beta^{S_1}_{g,k} \approx \beta^{S_2}_{h,l}} \lambda_S(\beta^{S_1}_{g,k} - \beta^{S_2}_{h,l})^2
\end{equation}

@@

When there is a one-to-one orthology between the species being considered (ie different cell-lines of the same organism), the choice of $\lambda_S$ allows one to interpolate between fitting each network independently ($\lambda_S=0$) and pooling data together as if it came from one source ($\lambda_S=\inf$). Part of the appeal of the approach, however, is that it allows pooling of data even when there is incomplete orthology. By introducing constraints on the similarity of individual interactions, rather than on the networks as a whole [CITE IBM?], we can pool some information across species even when an arbitrarily small fraction of genes have orthologs. This is particularly useful when dealing with a large number of species; pairwise orthologies may be nearly complete even when the number of genes present in every organism is small.

\subsection{Solving}
The problem of solving a multi-valued regression problem can be converted into the problem of solving a single valued regression problem by vectorization.
Specifically, if $Y \in R^{N \times M}$ then we form $A =$ diagonally concatenate $X$, $M$ times (with the remaining entries 0),
$y =$ vertically concatenate the columns of $Y$

$$\underset{w}{min} ||Aw - y||_2^{~2}$$

We can then define $W_{i,j} = \beta_{j \times N + i}$ which is the desired form.
This is equivalent to the original problem due to the block structure of matrix multiplication

Given $X_1... X_S$ matrices, $X_i \in R^{N_i \times K_i}$ we define $N = \sum_{i=0}^S N_i$, $K = \sum_{i=0}^SK_i$ and $X \in R^{N \times K}$ by diagonally concatenating $X_1...X_S$.
In this new matrix, the first $K_1$ features correspond to features of $X_1$, the next $K_2$ features the features of $X_2$ etc.
Similarly, the first $N_1$ samples correspond to samples of $X_1$, the next $N_2$ to $X_2$ etc.
Solving the regression problem associated with each source is equivalent to solving the problem $\underset{W}{argmin}||XW-Y||$ where $Y$ is defined in the same way as $X$.
The weight matrix $W$ has a block diagonal structure:

\begin{figure}[h!]
    \centering
    \includegraphics[scale=.5,trim=0mm 0mm 0mm 0mm,clip]{direct_solve.eps}
\end{figure}

The fused multi-source regression problem can solved by constructing a multi-source design matrix $X$ and data matrix $Y$, and then vectorizing.
The regression problems associated with two columns $g,h$ of $Y$, where $g$ came from $Y_G$ and $h$ came from $Y_H$, are independent, unless there is a constraint chain of length $L$ with the form (where $x$ is arbitrary)

%$$(G, j_0, g, J_1, j_1, m_1) \rightarrow (J_1, j_1, m_1, J_2, j_2, m_2) \rightarrow (J_2,j_2,m_2,J_3,j_3,_m3)...(J_P, j_P, m_P, H, j_{P+1}, h)$$
$$(G, j_0, g, J_1, j_1, m_1) \rightarrow (J_1, x, m_1, J_2, j_2, m_2)$$

We use depth-first search to identify linked columns of $Y$, and then solve the fused regression problem directly by vectorizing those columns.
In order to incorporate the fusion penalty, we append to the resulting design matrix one row \emph{per connected constraint}, which contains $\lambda_S$ in the column associated with $G,i,j,$ and $-\lambda_S$ to the column associated with $J, i',j'$ for constraint $(G,i,j,J,i',j')$. Call this resulting matrix $X_P$.
The vector $Y$ is augmented by appending an appropriate number of zeros to produce $Y_P$. As a result, each entry in the output of $X_P b$ in the columns associated with each constraint contain $b_{i,j} - b_{i',j,i}$.
These entries each contribute  $||b_{i,j} - b_{i',\{ji\}}||^{~2}$ to the squared error $||X_Pb - Y_P||_2^{~2}$

In the context of regression for GRN inference with multiple cell sources, each gene $j$ from source $1$ has constraints $\forall_{i \in \{TFs\}} (1, i, j, 2, i’, j’)$ where $i',j'$ are the same gene in a different cell line.
As a result, each gene has only one constraint and $L=2$.
%In general, however, solving the fused regression problem directly, even after factoring constraints, involves at least one inversion of an $L\sum_{i=0}^S K_i \times L \sum_{i=0}^S K_i$ matrix, taking $O((L\sum_{i=0}^S K_i)^3)$.
In general, however, solving the fused regression problem directly, even after factoring constraints, involves at least one inversion of an $p\sum_{i=0}^S K_i \times p \sum_{i=0}^S K_i$ matrix where $p$ is the size of a column subset such that there is a constraint chain $L$ linking an element of each column to an element in every other column in that subset, taking $O((p\sum_{i=0}^S K_i)^3)$.
Although the sparse structure of $X_P$ makes this somewhat more tractable, it is still impractical for large $L$.

\subsubsection{Equivalent prior}
Our L2 fusion penalty is equivalent to assuming a Gaussian prior with variance proportional to $\frac{1}{\lambda_S}$ on differences in parameters with fusion constraints. Combined with the regularization constraints, this forms a multivariate Gaussian prior on weights in both networks. When using our method on systems with incomplete orthology mapping, this results in different prior probability distributions for those parameters with fusion constraints and those without. We adjust for this by solving for a constant to multiply fused interactions by to equalize the volumes of the prior distributions. The intent of this adjustment is to ensure that constrained interactions are not on average more highly penalized, which may tend to drive their weight towards zero, causing them to be excluded from the network.
